---
title: "New York county analysis"
author: "Sandhya Kajeepeta"
date: "2024-07-17"
output: html_document
---

```{r load packages,echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(sf)
library(readxl)
library(MASS)
library(pscl)
```


```{r load data, echo=FALSE}
df2 <- read_xlsx("./County eviction data.xlsx")
#create variables of interest
df2$res_evic_rate <- df2$`Residential Evictions`/df2$`Total Renter-occupied housing units`*100
df2$res_holdover_evic_rate <- df2$`Residential Holdover Evictions`/df2$`Total Renter-occupied housing units`*100

#rename vars
df2$res_evics <- df2$`Residential Evictions`
df2$res_holdover_evics <- df2$`Residential Holdover Evictions`
df2$rental_units <- df2$`Total Renter-occupied housing units`

df2$perc_black_rent <- df2$`Black Renter-occupied housing units`/df2$`Total Renter-occupied housing units`*10
```

```{r plot vars, echo=FALSE, results=FALSE, warning=FALSE, fig.show='hide'}
plot <-ggplot(df2, mapping=aes(x=perc_black_rent, y=res_holdover_evic_rate)) + geom_point()
plot

plot <-ggplot(df2, mapping=aes(x=perc_black_rent, y=res_holdover_evics/res_evics)) + geom_point()
plot

hist(df2$res_evic_rate)
hist(df2$perc_black_rent)
```

```{r calc perc holdover, echo=FALSE, warning=FALSE}
perc_h <- df2 %>% dplyr::select(res_holdover_evics, res_evics) %>% summarise_all(sum, na.rm=T)
perc_h$res_holdover_evics/perc_h$res_evics
```
Nearly 20% of all residential evictions in New York State in 2022 were holdover evictions.


```{r poisson or neg binomial, echo=FALSE, results=FALSE, fig.show='hide'}
p_model <- glm(formula=res_evics~perc_black_rent + offset(log(rental_units)), data=df2, family="poisson")
nb_model <- glm.nb(formula=res_evics~perc_black_rent + offset(log(rental_units)), data=df2)

#Residual plot for Poisson regression
p_res <- resid(p_model)
plot(fitted(p_model), p_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Poisson')
abline(0,0)

#Residual plot for negative binomial regression 
nb_res <- resid(nb_model)
plot(fitted(nb_model), nb_res, col='steelblue', pch=16,
     xlab='Predicted Offers', ylab='Standardized Residuals', main='Negative Binomial')
abline(0,0)

#Residuals for negative binomial plot are much smaller, Poisson residuals extend all the way to 60 - suggests that negative binomial is a better fit than Poisson

pchisq(2 * (logLik(nb_model) - logLik(p_model)), df = 1, lower.tail = FALSE)

#Negative binomial offers significantly better fit
```

```{r analysis, echo=FALSE, warning=FALSE, message=FALSE}
m <- glm.nb(formula=res_evics~perc_black_rent + offset(log(rental_units)), data=df2)
print("Residential eviction filing rate")
est <- exp(cbind(Estimate = coef(m), confint(m)))
kable(est, "pipe")

m <- glm.nb(formula=res_holdover_evics~perc_black_rent + offset(log(rental_units)), data=df2)
print("Residential holdover eviction filing rate")
est <- exp(cbind(Estimate = coef(m), confint(m)))
kable(est, "pipe")

```

New York counties with a higher percentage of Black renters face a higher rate of total residential eviction filings and residential holdover eviction filings.

A 10-percentage-point increase in the percent of Black renters in a county is associated with a 55% increase in the rate of total residential eviction filings.

A 10-percentage-point increase in the percent of Black renters in a county is associated with a 33% increase in the rate of residential holdover eviction filings.

*Sources: https://app.powerbigov.us/view?r=eyJrIjoiZGE3NzljYmItYTBmZC00OGI2LTliYTgtYzY5ZjI0N2U0MWYxIiwidCI6IjM0NTZmZTkyLWNiZDEtNDA2ZC1iNWEzLTUzNjRiZWMwYTgzMyJ9 (2022); https://data.census.gov/table/ACSST5Y2021.S2502?q=S2502&g=040XX00US36,36$0500000&tp=true (2021)*